Randomized trace estimation of parameter-dependent matrices applied to spectral density approximation

### 2024-05-23 paper/main.tex
start writing introduction

%\todo{
%\begin{itemize}
%    \item Unify index notation
%    \item Double-check proofs
%    \item Add all references
%    \item Run numerical experiments 
%    \item Write introduction, abstract, and discussion
%\end{itemize}
%}

@@

\begin{itemize}
    \item Definition of parameter-dependent matrix ($\mtx{B}(t) \in \mathbb{R}^{n \times n}$, every entry depends continuously on $t \in [a,b]$ \cite{kressner-2023-randomized-lowrank})
    \item Trace estimation through mat-vecs (if matrix is only accessed implicitly [Hessian, Graph] or for matrix functions [partition function, spectral density, exponential integrators])
    \item Mention existing trace estimation methods for constant matrices (quadratic [Hutchinson, Girard], Hutch++ \cite{meyer-2021-hutch-optimal}, Nyström++ \cite{persson-2022-improved-variants}, XTrace [Tropp], Krylov aware [Chen]...)
\end{itemize}

\paragraph{Spectral density approximation}
\begin{itemize}
    \item Definition of spectral density ($\phi(t) = n^{-1}\sum_{i=1}^{m} \delta(t - \lambda_i)$ for matrix $\mtx{A}$ with eigenvalues $\lambda_1, \dots, \lambda_n$)
    \item Applications of spectral density approximation (electronic structure calculations \cite{lin-2017-randomized-estimation}, neural network optimization [Hessian, ...], graph processing [filtering, ...])
    \item Link to trace estimation of a matrix function (smoothing with Gaussian \cite{lin-2017-randomized-estimation} or Haydock [Haydock, Lin] kernel)
    \item Existing theory for spectral density approximation (Wasserstein-1 error [Braverman, Chen], missing $1/\varepsilon$ guarantees)
\end{itemize}

\paragraph{Contribution}
\begin{itemize}
    \item Development of a unified family of estimators for many evaluations of parameter, i.e. try to keep complexity with each evaluation as low as possible
    \item Extends \cite{kressner-2023-randomized-lowrank} to trace estimation
    \item Provides theoretical guarantees for methods in \cite{lin-2017-randomized-estimation}
\end{itemize}

\paragraph{Notation}
We follow the notation and conventions which are used in contemporary literature in this field. Further,
\begin{itemize}
    \item we denote with $\mathbb{N}$ all positive whole numbers and with $\mathbb{R}$ all real numbers;
    \item $e$ denotes Euler's number and $\log$ is the corresponding logartihm;
    \item the ceiling function $\lceil x \rceil$ of a real number $x$ returns the greatest integer larger than or equal to $x$;
    \item scalars are represented by lower case Greek or Latin letters ($s$, $\varepsilon$, \dots),
    vectors are additionally printed in bold ($\vct{v}$, $\vct{\psi}$, \dots),
    and matrices are additionally capitalized ($\mtx{A}$, $\mtx{\Omega}$, \dots);
    \item the expected value of a random variable $x$ is denoted with $\mathbb{E}[x]$. The $k$-th moment of a random variable $x$ is defined as $\mathbb{E}^{k}[x] = \left(\mathbb{E}\left[ | x |^{k} \right] \right)^{\frac{1}{k}}$;
    \item the identity matrix $\mtx{I}_n = \diag(1, \dots, 1) \in \mathbb{R}^{n \times n}$
    carries ones on its diagonal and zeros everywhere else. The zero matrix
    $\mtx{0}_{n \times m} \in \mathbb{R}^{n \times m}$ consists of only zero entries;
    \item in a Gaussian random vector and a Gaussian random matrix, all entries are independently sampled from the standard normal distribution;
    \item matrix norms are denoted with $\lVert \cdot \rVert _{\cdot}$. For symmetric matrices, the $p$-Schatten norm is
    $\lVert \mtx{A} \rVert _{(p)} = (\sum_{i=1}^n |\lambda_i|^p)^{1/p}$. We refer to the spectral norm as $\lVert \mtx{A} \rVert _2 = \lVert \mtx{A} \rVert _{(\infty)} = \max_{i=1,\dots,n}|\lambda_i|$, the nuclear norm as $\lVert \mtx{A} \rVert _{\ast} = \lVert \mtx{A} \rVert _{(1)} = \sum_{i=1}^{n} |\lambda_i|$, and the Frobenius norm as $\lVert \mtx{A} \rVert _F = \lVert \mtx{A} \rVert _{(2)} = (\sum_{i=1}^{n} |\lambda_i|^2)^{\frac{1}{2}}$.
\end{itemize}

\section{Analysis of parameter-dependent trace estimators}
\label{sec:analysis}

@@

\paragraph{Quadratic trace estimator} If we can only access the parameter-dependent matrix through matrix-vector products, we can estimate the trace with the quadratic trace estimator \cite{}: We take $n_{\Psi} \in \mathbb{N}$ Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\psi}} \in \mathbb{R}^{n}$ to form
\begin{equation}
    \Hutch{n_{\Psi}}(\mtx{A}(t)) = \frac{1}{n_{\Psi}} \sum_{j=1}^{n_{\Psi}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j.
    \label{equ:quadratic-trace-estimator}
\end{equation}
If we gather the random vectors in a matrix $\mtx{\Psi} = [\vct{\psi}_1 ~ \cdots ~ \vct{\psi}_{n_{\psi}}] \in \mathbb{R}^{n \times n_{\Psi}}$ we may rewrite $\Hutch{n_{\Psi}}(\mtx{A}(t)) = n_{\Psi}^{-1} \Trace( \mtx{\Psi}^{\top} \mtx{A}(t) \mtx{\Psi})$. Other choices for the distribution of the entries of the random vectors are possible \cite{}. However, our theoretical developments only hold in the Gaussian case.

\paragraph{Nyström trace approximation} Alternatively, the trace of a symmetric matrix whose rank is significantly lower than its size can be approximated well by using a Gaussian sketching matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\Omega}}$ to form the Nyström approximation
\begin{equation}
    \widehat{\mtx{A}}(t) = (\mtx{A}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{A}(t) \mtx{\Omega})^{\dagger} (\mtx{A}(t) \mtx{\Omega})^{\top}.
\end{equation}
Then we define the trace estimator $\Nystr{n_{\Omega}}(\mtx{A}(t)) = \Trace( \widehat{\mtx{A}}(t) )$.
Thanks to the invariance of the trace under cyclic permutation of its arguments and the symmetry of the matrix, we may also rewrite
\begin{equation}
    \Nystr{n_{\Omega}}(\mtx{A}(t)) = \Trace( (\mtx{\Omega}^{\top} \mtx{A}(t) \mtx{\Omega})^{\dagger} ( \mtx{\Omega}^{\top} \mtx{A}(t)^2 \mtx{\Omega})).
    \label{equ:nystrom-trace-estimator}
\end{equation}

\paragraph{Nyström++ estimator} Finally, a combination of the two estimators yields an analogue for the Nyström++ estimator \cite{persson-2022-improved-variants} in the parameter-dependent setting
\begin{equation}
    \Nystrpp{n_{\Omega}}{n_{\Psi}}(\mtx{A}(t)) = \Nystr{n_{\Omega}}(\mtx{A}(t)) + \Hutch{n_{\Psi}}(\mtx{A}(t) - \widehat{\mtx{A}}(t)).
    \label{equ:nystrompp-trace-estimator}
\end{equation}
It is understood that $\widehat{\mtx{A}}$ is computed with the same sketching matrix as the one used for the trace estimator $\Nystr{n_{\Omega}}$. We can interpret this estimator as an interpolation between the trace of the Nyström approximation and the quadratic trace estimator.

\paragraph{$L^{1}$ norm}
\todo{Give definition of $L^1$ norm and reason for choosing it}

\subsection{Parameter-dependent quadratic trace estimator}
\label{subsec:quadratic}

We first consider \refequ{equ:quadratic-trace-estimator}.
\todo{Discuss theory for constant matrix
\begin{equation}
    \lVert f \rVert _1 = \int_{a}^{b} | f(t) | \mathrm{d}t
\end{equation}
}

\begin{theorem}{Parameter-dependent quadratic trace estimator}{quadratic}
    Let $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ continuously depend on $t \in [a, b]$. Then for any $k \in \mathbb{N}$ and $\gamma \geq 1$ it holds with probability at least $1 - \gamma^{-2k}$
    \begin{equation}
        \int_{a}^{b} \left| \Trace(\mtx{A}(t)) - \Hutch{n_{\Psi}}(\mtx{A}(t)) \right| ~\mathrm{d}t < 16 \gamma k \frac{1}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F  ~\mathrm{d}t.
    \end{equation}
    In particular, if we choose $n_{\Psi} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$, then with probability at least $1-\delta$ for $\delta \in (0, e^{\frac{1}{2}})$ and any $\varepsilon > 0$ we have $\int_{a}^{b} | \Trace(\mtx{A}(t)) - \Hutch{n_{\Psi}}(\mtx{A}(t)) | ~\mathrm{d}t < \varepsilon \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t$.
\end{theorem}

The idea behind the proof is to bound the higher order moments for fixed parameter values and then use Markov's inequality to transition to the continuous parameter case. We first only consider the $1$-query estimator and subsequently use a trick to carry the result over to the general case.
\begin{proof} 
    We define the 1-query quadratic trace estimate 
    \begin{equation}
        r(t) = \vct{\psi}^{\top} \mtx{B}(t) \vct{\psi} - \Trace(\mtx{B}(t))
    \end{equation}
    for a parameter-dependent matrix $\mtx{B}(t)$ over the real numbers with the Gaussian random vector $\vct{\psi}$ in a fixed $t \in [a,b]$.
    From the proof of \cite[Lemma 3]{cortinovis-2022-randomized-trace} we know that $r(t)$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B}(t) \rVert _F^2, 2 \lVert \mtx{B}(t) \rVert _2)$. Thus, by \cite[Theorem 2.3]{boucheron-2013-basic-inequalities} this implies that for every $k \in \mathbb{N}$
    \begin{align}
        \mathbb{E}\left[ r(t)^{2 k} \right]
        &\stackrel{\text{\cite{boucheron-2013-basic-inequalities}}}{\leq} k! \left( 16 \lVert \mtx{B}(t) \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B}(t) \rVert _2 \right)^{2 k} \notag \\
        &= k! 2^{4 k} \lVert \mtx{B}(t) \rVert _F^{2 k} + (2 k)! 2^{6 k} \lVert \mtx{B}(t) \rVert _2^{2 k}.
    \end{align}
    Since $\lVert \mtx{B}(t) \rVert _2 \leq \lVert \mtx{B}(t) \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$ we can upper bound 
    \begin{equation}
        \mathbb{E}\left[ r(t)^{2 k} \right] \leq \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B}(t) \rVert _F^{2k}.
    \end{equation}
    Stirling's approximation \cite{robbins-1955-remark-stirling} bonuds $(2 k)! < 2 \sqrt{\pi k}  e^{\frac{1}{24 k}} ( 2 k / e )^{2 k}$. Consequently, the moments of $r(t)$ for a fixed $t$ are limited by
    \begin{equation}
        \mathbb{E}^{2k}\left[ r(t) \right]
        < \left( \frac{9}{4} \sqrt{\pi k} e^{\frac{1}{24 k}} \right)^{\frac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B}(t) \rVert _F < 16 k \lVert \mtx{B}(t) \rVert _F,
        \label{equ:quadratic-trace-onequery-fixed}
    \end{equation}
    since it can be checked that $(\frac{9}{4} \sqrt{\pi k} e^{\frac{1}{24 k}})^{\frac{1}{2k}} < e$, due to the monotonous decrease of this expression in $k \in \mathbb{N}$.

    Now we make the transition to the continuous. Minkowski's integral inequality \cite[Theorem 2.2]{hardy-1952-inequalities} allows us to apply \refequ{equ:quadratic-trace-onequery-fixed} in the continuous setting: 
    \begin{equation}
        \mathbb{E}^{2 k}\left[ \int_{a}^{b} |r(t)|~\mathrm{d}t  \right]
        \stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}\left[ |r(t)|^{2 k}\right]^{\frac{1}{2 k}}~\mathrm{d}t
        < 16 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
    \end{equation}
    By Markov's inequality \cite{}, for any $k \in \mathbb{N}$ and $\gamma \geq 1$, it holds with probability at least $1 - \gamma^{-2 k}$
    \begin{equation}
        \int_{a}^{b} |r(t)|~\mathrm{d}t < 16 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
        \label{equ:quadratic-trace-onequery-uniform}
    \end{equation}

    To extend the bound to the $n_{\Psi}$-query quadratic trace estimator, we use a technique from the proof of \cite[Proof of Theorem 1]{cortinovis-2022-randomized-trace}. Let
    \begin{equation}
        \mtx{B}(t)
        = \frac{1}{n_{\Psi}} \begin{pmatrix}
            \mtx{A}(t) & & \\
            & \ddots & \\
            & & \mtx{A}(t)
        \end{pmatrix}
        \quad \text{and} \quad
        \vct{\psi} = \begin{pmatrix}
            \vct{\psi}_1 \\
            \vdots \\
            \vct{\psi}_{n_{\Psi}}
        \end{pmatrix}
    \end{equation}
    where $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\Psi}} \in \mathbb{R}^{n}$ are independent Gaussian random vectors. Then $\lVert \mtx{B}(t) \rVert _F = \frac{1}{\sqrt{n_{\Psi}}} \lVert \mtx{A}(t) \rVert _F$ and
    \begin{equation}
        r(t) = \frac{1}{n_{\Psi}} \sum_{j=1}^{n_{\Psi}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j - \Trace(\mtx{A}(t)).
    \end{equation}
    Hence, we conclude from \refequ{equ:quadratic-trace-onequery-uniform} that for any $k \in \mathbb{N}$ and $\gamma \geq 1$ with probability at least $1 - \gamma^{-2 k}$
    \begin{equation}
        \int_{a}^{b} \left| \frac{1}{n_{\Psi}} \sum_{j=1}^{n_{\Psi}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j - \Trace(\mtx{A}(t)) \right| ~ \mathrm{d}t
        < 16 k \gamma \frac{1}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t.
    \end{equation}

    Setting $\delta = \gamma^{-2 k}$ and choosing $k = \lceil \log(\delta^{-1}) \rceil$ we get that for all $\delta \in (0, e^{-\frac{1}{2}})$
    \begin{equation}
        k \gamma = \lceil \log(\delta^{-1}) \rceil \delta^{-\frac{1}{2\lceil \log(\delta^{-1}) \rceil}}
        = \lceil \log(\delta^{-1}) \rceil e^{\frac{1}{2}\frac{\log(\delta^{-1})}{\lceil \log(\delta^{-1}) \rceil}}
        \leq 2 \log(\delta^{-1}) e^{\frac{1}{2}},
    \end{equation}
    from which follows the second part of the theorem.
    %Consequently, with probability $1 - \delta$ we have
    %\begin{equation}
    %    \int_{a}^{b} \left| \frac{1}{n_{\Psi}} \sum_{i=1}^{n_{\Psi}} \vct{\psi}_i^{\top} \mtx{A}(t) \vct{\psi}_i - \Trace(\mtx{A}(t)) \right| ~ \mathrm{d}t \leq 32 e^{\frac{1}{2}} \frac{\log(\delta^{-1})}{\sqrt{n_{\Psi}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t
    %\end{equation}
\end{proof}

>>>>

%\todo{
%\begin{itemize}
%    \item Unify index notation
%    \item Double-check proofs
%    \item Add all references
%    \item Run numerical experiments 
%    \item Write introduction, abstract, and discussion
%    \item ArXiv HTML support (check if packages)
%\end{itemize}
%}

\todo{
    \tableofcontents
}

@@

In numerous problems across physics, chemistry, engineering, and data science, eigenvalues and especially their distribution are crucial for understanding the properties of a system: In electronic structure calculations they represent the energy levels which electrons occupy \cite{lin-2017-randomized-estimation, drabold-1993-maximum-entropy, ducastelle-1970-moments-developments, haydock-1972-electronic-structure}, in neural network optimization they are indicative of the topology of the loss landscape \cite{yao-2020-pyhessian-neural, ghorbani-2019-investigation-neural}, and in graph processing they can uncover hidden graph motifs \cite{huang-2021-density-states}. The distribution of the eigenvalues $\lambda_1, \dots, \lambda_n \in \mathbb{R}$ of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ can be captured with the spectral density
\begin{equation}
    \phi(t) = \frac{1}{n} \sum_{i=1}^{n} \delta(t - \lambda_i).
\end{equation}
Clearly, assembling this expression amounts to computing all eigenvalues of the matrix; an operation which is often prohibitively expensive. Therefore, many techniques have been proposed for approximating $\phi$. For example, by matching its moments with a discrete density function \cite{cohen-steiner-2018-approximating-spectrum, braverman-2022-sublinear-time}, by deriving a quadrature from the Lanczos procedure \cite{lin-2016-approximating-spectral, chen-2021-analysis-stochastic}, or by explicit polynomial expansion \cite{weisse-2006-kernel-polynomial, lin-2016-approximating-spectral}. In many applications, it is also acceptable to trade in some of the finer details of the spectrum for a better approximability by smoothing $\phi$, typically through convolution with a Gaussian \cite{lin-2016-approximating-spectral, lin-2017-randomized-estimation} or a Lorentzian \cite{haydock-1972-electronic-structure, lin-2016-approximating-spectral}. This converts the problem to estimating the trace of a smooth parameter-dependent matrix function.

\todo{
\begin{itemize}
    \item Definition of parameter-dependent matrix ($\mtx{B}(t) \in \mathbb{R}^{n \times n}$, every entry depends continuously on $t \in [a,b]$ \cite{kressner-2023-randomized-lowrank})
    \item Trace estimation through mat-vecs (if matrix is only accessed implicitly \cite{pearlmutter-1994-fast-exact} or for complicated matrix functions [partition function, spectral density, exponential integrators])
    \item Existing trace estimation methods for constant matrices: Girard-Hutchinson \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, variance reduced versions thereof \cite{gambhir-2017-deflation-method, saibaba-2017-randomized-matrixfree, lin-2017-randomized-estimation, meyer-2021-hutch-optimal, persson-2022-improved-variants, epperly-2024-xtrace-making} which have only recently been specialized to trace estimation of matrix functions \cite{chen-2023-krylovaware-stochastic}
    \item Trace estimation for parameter-dependent matrices (Nyström++ but no analysis \cite{lin-2017-randomized-estimation}, low-rank approximation \cite{kressner-2023-randomized-lowrank}, dynamic trace estimation but only discrete setting [Woodruff, Dharangutte])
    %\item Error analysis for spectral density approximation (Wasserstein-1 error [Braverman, Chen], missing $1/\varepsilon$ guarantees)
\end{itemize}

\paragraph{Contribution}
\begin{itemize}
    \item Theoretical guarantees in the $L^1$-norm for three trace estimators on parameter-dependent matrices (Girard-Hutchinson, Nyström, Nyström++), in particular, $\mathcal{O}(\varepsilon^{-1})$ scaling with number of mat-vecs
    \item A simple and efficient algorithm for estimating spectral densities using any of the three estimators using Chebyshev expansion, including theoretical guarantees
\end{itemize}
}

\paragraph{Notation}
We follow the notation and conventions which are used in contemporary literature in this field. Further,
\begin{itemize}
    \item we denote with $\mathbb{N}$ all positive whole numbers and with $\mathbb{R}$ all real numbers;
    \item $e$ denotes Euler's number and $\log$ is the corresponding logartihm;
    \item the ceiling function $\lceil x \rceil$ of a real number $x$ returns the greatest integer larger than or equal to $x$;
    \item scalars are represented by lower case Greek or Latin letters ($s$, $\varepsilon$, \dots),
    vectors are additionally printed in bold ($\vct{v}$, $\vct{\psi}$, \dots),
    and matrices are additionally capitalized ($\mtx{A}$, $\mtx{\Omega}$, \dots);
    \item the expected value of a random variable $x$ is denoted with $\mathbb{E}[x]$. The $k$-th moment of a random variable $x$ is defined as $\mathbb{E}^{k}[x] = \left(\mathbb{E}\left[ | x |^{k} \right] \right)^{\sfrac{1}{k}}$;
    \item the identity matrix $\mtx{I}_n = \diag(1, \dots, 1) \in \mathbb{R}^{n \times n}$
    carries ones on its diagonal and zeros everywhere else. The zero matrix
    $\mtx{0}_{n \times m} \in \mathbb{R}^{n \times m}$ consists of only zero entries;
    \item in a Gaussian random vector and a Gaussian random matrix, all entries are independently sampled from the standard normal distribution;
    \item matrix norms are denoted with $\lVert \cdot \rVert$. For symmetric matrices, the $p$-Schatten norm is
    $\lVert \mtx{A} \rVert _{(p)} = (\sum_{i=1}^n |\lambda_i|^p)^{1/p}$. We refer to the spectral norm as $\lVert \mtx{A} \rVert _2 = \lVert \mtx{A} \rVert _{(\infty)} = \max_{i=1,\dots,n}|\lambda_i|$, the nuclear norm as $\lVert \mtx{A} \rVert _{\ast} = \lVert \mtx{A} \rVert _{(1)} = \sum_{i=1}^{n} |\lambda_i|$, and the Frobenius norm as $\lVert \mtx{A} \rVert _F = \lVert \mtx{A} \rVert _{(2)} = (\sum_{i=1}^{n} |\lambda_i|^2)^{\sfrac{1}{2}}$.
\end{itemize}

\todo{
\paragraph{Reproducibility} Fully reproducible paper hosted on GitHub.
}
\section{Analysis of trace estimators for parameter-dependent matrices}
\label{sec:analysis}

@@

\paragraph{Girard-Hutchinson estimator} If we can only access the parameter-dependent matrix through matrix-vector products, we can estimate the trace with the Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator}: We take $n_{\mtx{\Psi}}$ Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ to form
\begin{equation}
    \Hutch{\mtx{\Psi}}(\mtx{A}(t))
    = \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j
    = \frac{1}{n_{\mtx{\Psi}}} \Trace( \mtx{\Psi}^{\top} \mtx{A}(t) \mtx{\Psi})
    \label{equ:hutchinson-trace-estimator}
\end{equation}
where $\mtx{\Psi} = [\vct{\psi}_1 ~ \cdots ~ \vct{\psi}_{n_{\mtx{\Psi}}}] \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$. Other choices for the distribution of the entries of the random vectors are possible, for example by uniformly sampling from $\{-1, +1\}$ or from the $(n-1)$-sphere. However, our theoretical developments only hold in the Gaussian case.

\paragraph{Nyström approximation} Alternatively, the trace of a symmetric matrix whose rank is significantly lower than its size can be approximated well by using a Gaussian sketching matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ to form the Nyström approximation \cite{gittens-2013-revisiting-nystrom}
\begin{equation}
    \Nystr{\mtx{\Omega}}{\mtx{A}}(t) = (\mtx{A}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{A}(t) \mtx{\Omega})^{\dagger} (\mtx{A}(t) \mtx{\Omega})^{\top}.
\end{equation}
Then we use the trace estimator $\Trace(\Nystr{\mtx{\Omega}}{\mtx{A}}(t))$. Thanks to the invariance of the trace under cyclic permutation of its arguments and the symmetry of the matrix, we may also rewrite
\begin{equation}
    \Trace(\Nystr{\mtx{\Omega}}{\mtx{A}}(t)) = \Trace( (\mtx{\Omega}^{\top} \mtx{A}(t) \mtx{\Omega})^{\dagger} ( \mtx{\Omega}^{\top} \mtx{A}(t)^2 \mtx{\Omega})).
    \label{equ:nystrom-trace-estimator}
\end{equation}

\paragraph{Nyström++ estimator} Finally, a combination of the two estimators yields an analogue for the Nyström++ estimator \cite{persson-2022-improved-variants} in the parameter-dependent setting
\begin{equation}
    \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{A}(t)) = \Trace(\Nystr{\mtx{\Omega}}{\mtx{A}}(t)) + \Hutch{\mtx{\Psi}}(\mtx{A}(t) - \Nystr{\mtx{\Omega}}{\mtx{A}}(t)),
    \label{equ:nystrompp-trace-estimator}
\end{equation}
which has originally been proposed in \cite{lin-2017-randomized-estimation}. %We can interpret this estimator as an interpolation between the trace of the Nyström approximation and the Girard-Hutchinson estimator.

\paragraph{$L^{1}$ norm}
We define the $L^{1}$ norm of a square-integrable function $f:[a,b] \to \mathbb{R}$ as
\begin{equation}
    \int_{a}^{b} | f(t) | \mathrm{d}t.
\end{equation}
\todo{Motivate why we use this.}

\subsection{Girard-Hutchinson estimator for parameter-dependent matrices}
\label{subsec:hutchinson}

We first consider \refequ{equ:hutchinson-trace-estimator}.
\todo{Discuss theory for constant matrix
}

\begin{theorem}{Girard-Hutchinson estimator for parameter-dependent matrices}{hutchinson}
    Let $\mtx{A}(t) \in \mathbb{R}^{n \times n}$ continuously depend on $t \in [a, b]$. Then for any $k \in \mathbb{N}$ and $\gamma \geq 1$ it holds with probability at least $1 - \gamma^{-2k}$
    \begin{equation}
        \int_{a}^{b} \left| \Trace(\mtx{A}(t)) - \Hutch{\mtx{\Psi}}(\mtx{A}(t)) \right| ~\mathrm{d}t < 16 \gamma k \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F  ~\mathrm{d}t.
    \end{equation}
    In particular, if we choose $n_{\mtx{\Psi}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})^2)$, then with probability at least $1-\delta$ for $\delta \in (0, e^{\sfrac{1}{2}})$ and any $\varepsilon > 0$ we have $\int_{a}^{b} | \Trace(\mtx{A}(t)) - \Hutch{\mtx{\Psi}}(\mtx{A}(t)) | ~\mathrm{d}t < \varepsilon \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t$.
\end{theorem}

The idea behind the proof is to bound the higher order moments for fixed parameter values and then use Markov's inequality to transition to the continuous parameter case. We first only consider the $1$-query estimator and subsequently use a trick to carry the result over to the general case.
\begin{proof} 
    We define the 1-query quadratic trace estimate 
    \begin{equation}
        r(t) = \vct{\psi}^{\top} \mtx{B}(t) \vct{\psi} - \Trace(\mtx{B}(t))
    \end{equation}
    for a parameter-dependent matrix $\mtx{B}(t)$ over the real numbers with the Gaussian random vector $\vct{\psi}$.

    First, we consider $r(t)$ for a fixed $t \in [a,b]$ and therefore temporarily ignore the parameter-dependence. From the proof of \cite[Lemma 3]{cortinovis-2022-randomized-trace} we know that $r$ is sub-Gamma with parameters $(v, c) = (2 \lVert \mtx{B} \rVert _F^2, 2 \lVert \mtx{B} \rVert _2)$. Thus, by \cite[Theorem 2.3]{boucheron-2013-basic-inequalities} this implies that for every $k \in \mathbb{N}$
    \begin{align}
        \mathbb{E}\left[ r^{2 k} \right]
        &\stackrel{\text{\cite{boucheron-2013-basic-inequalities}}}{\leq} k! \left( 16 \lVert \mtx{B} \rVert _F^2 \right)^k + (2 k)! \left( 8 \lVert \mtx{B} \rVert _2 \right)^{2 k} \notag \\
        &= k! 2^{4 k} \lVert \mtx{B} \rVert _F^{2 k} + (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _2^{2 k}.
    \end{align}
    Since $\lVert \mtx{B} \rVert _2 \leq \lVert \mtx{B} \rVert _F$ and $k! 2^{4 k} + (2 k)! 2^{6 k} \leq \frac{9}{8}(2 k)! 2^{6 k}$ for any $k \in \mathbb{N}$ we can upper bound 
    \begin{equation}
        \mathbb{E}\left[ r^{2 k} \right] \leq \frac{9}{8} (2 k)! 2^{6 k} \lVert \mtx{B} \rVert _F^{2k}.
    \end{equation}
    Stirling's approximation \cite{robbins-1955-remark-stirling} bonuds $(2 k)! < 2 \sqrt{\pi k}  e^{\sfrac{1}{24 k}} ( 2 k / e )^{2 k}$. Consequently, the moments of $r$ for a fixed $t$ are limited by
    \begin{equation}
        \mathbb{E}^{2k}\left[ r \right]
        < \left( \frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}} \right)^{\sfrac{1}{2k}} \left( \frac{2 k}{e}\right) 2^{3} \lVert \mtx{B} \rVert _F < 16 k \lVert \mtx{B} \rVert _F,
        \label{equ:hutchinson-trace-onequery-fixed}
    \end{equation}
    since it can be checked that $(\frac{9}{4} \sqrt{\pi k} e^{\sfrac{1}{24 k}})^{\sfrac{1}{2k}} < e$, due to the monotonous decrease of this expression in $k \in \mathbb{N}$.

    Now we make the transition to the continuous. Minkowski's integral inequality \cite[Theorem 2.2]{hardy-1952-inequalities} allows us to apply \refequ{equ:hutchinson-trace-onequery-fixed} in the continuous setting: 
    \begin{equation}
        \mathbb{E}^{2 k}\left[ \int_{a}^{b} |r(t)|~\mathrm{d}t  \right]
        \stackrel{\text{\cite{hardy-1952-inequalities}}}{\leq} \int_{a}^{b} \mathbb{E}\left[ |r(t)|^{2 k}\right]^{\sfrac{1}{2 k}}~\mathrm{d}t
        < 16 k \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
    \end{equation}
    By Markov's inequality, for any $k \in \mathbb{N}$ and $\gamma \geq 1$, it holds with probability at least $1 - \gamma^{-2 k}$
    \begin{equation}
        \int_{a}^{b} |r(t)|~\mathrm{d}t < 16 k \gamma \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t.
        \label{equ:hutchinson-trace-onequery-uniform}
    \end{equation}

    To extend the bound to the $n_{\mtx{\Psi}}$-query quadratic trace estimator, we use a technique from the proof of \cite[Theorem 1]{cortinovis-2022-randomized-trace}. Let
    \begin{equation}
        \mtx{B}(t)
        = \frac{1}{n_{\mtx{\Psi}}} \begin{pmatrix}
            \mtx{A}(t) & & \\
            & \ddots & \\
            & & \mtx{A}(t)
        \end{pmatrix}
        \quad \text{and} \quad
        \vct{\psi} = \begin{pmatrix}
            \vct{\psi}_1 \\
            \vdots \\
            \vct{\psi}_{n_{\mtx{\Psi}}}
        \end{pmatrix}
    \end{equation}
    where $\vct{\psi}_1, \dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$ are independent Gaussian random vectors. Then $\lVert \mtx{B}(t) \rVert _F = \lVert \mtx{A}(t) \rVert _F / \sqrt{n_{\mtx{\Psi}}}$ and
    \begin{equation}
        r(t) = \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j - \Trace(\mtx{A}(t)).
    \end{equation}
    Hence, we conclude from \refequ{equ:hutchinson-trace-onequery-uniform} that for any $k \in \mathbb{N}$ and $\gamma \geq 1$ with probability at least $1 - \gamma^{-2 k}$
    \begin{equation}
        \int_{a}^{b} \left| \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{A}(t) \vct{\psi}_j - \Trace(\mtx{A}(t)) \right| ~ \mathrm{d}t
        < 16 k \gamma \frac{1}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{A}(t) \rVert _F~\mathrm{d}t.
    \end{equation}
